Analysis of observational data, to calculate the Delayed-Tau model parameters

### 2025-03-09 thesis.qmd
uni-norm tau
turned tau to both uniform and normal prior in stan

the root $x_0\Leftrightarrow g(x_0) = 0$, can be found, for each galaxy. From @eq-A we can calculate the normilization constant of the galaxies, as well as their charachteristic timescales $\tau = t_{sf}\cdot x$.

@@

The second method for the calclulation of the parameters is the MCMC method. Using the logarithmic form of the model

@@

-   $\tau \sim \text{normal(mean=4 Gyr, }\sigma = 1 )$, within the limits $[1,20]$ Gyr,
-   $t_{sf}\sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$,
-   $\zeta \sim \text{normal(mean=1.3, }\sigma = 0.01)$, within the limits $[1,2]$,

and the likelihood:

-   $\log(\text{SFR}_{obs}) \sim \text{normal}\left(\log(\text{SFR}_\text{MCMC}), \sigma = 0.1\right)$

## According to pre-existing bibliography

### Problems with the method

## MCMC

>>>>

the root $x_0\Leftrightarrow g(x_0) = 0$, can be found, for each galaxy. From @eq-A we can calculate the normalization constant of the galaxies, as well as their characteristic timescales $\tau = t_{sf}\cdot x$.

@@

The second method for the calculation of the parameters is the MCMC method. Using the logarithmic form of the model

@@

-   $\tau \sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$,
-   $t_{sf}\sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$,
-   $\zeta \sim \text{normal(mean=1.3, }\sigma = 0.01)$, within the limits $[1,2]$,

and the likelihood $\log(\text{SFR}_{obs}) \sim \text{normal}\left(\log(\text{SFR}_\text{MCMC}), \sigma = 0.1\right)$, while taking the Stellar Masses of the galaxies as they are, without any variance, and $A_{del}$ is calculated by @eq-A. The MCMC run consists of 6 chains with 8000 iterations each (2500 iterations for the burn-in phase). Using this method, the galaxies can have different $t_{sf}$, while only taking positive values for their timescales.

The prior of the $\tau$ and $t_{sf}$ follow a uniform distribution, since we do not want to introduce any unwanted bias in the value selection of the MCMC sampler. While there is bibliography that could justify using a normal distribution for $\tau$, for example @speagleHighlyConsistentFramework2014 and @kroupaConstraintsStarFormation2020 explain that the galaxies of the Main Sequence are expected to have $3.5\lesssim\tau/\text{Gyr}\lesssim4.5$, this could significantly influence the results of galaxies with slow rising SFR's. In the same way a wrong prior for $t_{sf}$ could effect the results, especially of younger galaxies. While the choice of different priors is possible, for the aim of this thesis, the uniform distribution is the best choice.

Using these priors, the metrics of the Stan programm give good results ($\hat{R}<1.001$ and $\text{ESS} >1000$), while the calculated values correspond to the data of the catalog. Specifically, we use the data to recalculate the SFR from the predicted parameters, using the delayed-$\tau$ model and compare it to the observational data (@fig-comp-sfr-obs-pred), and also compare the normalization constant to the stellar masses of the galaxies (@fig-A-M-mcmc).

![Comparison of the observed SFR with the predicted by the parameters given by the MCMC run.](r_mcmc/plots/sfr_diff_plot.png){#fig-comp-sfr-obs-pred}

![Linear fit for $A_{del}\sim M_*$, with a correlation of 94%. The black dashed line is the linear regression weighted by $x$, with a fit $\log(A_{del}) = 0.95\cdot\log(M_*)+0.60, with a corralation of 98\%$](method_comparison/A_M_MCMC.png){#fig-A-M-mcmc}

In @fig-sSFR-A-both-methods, it can be seen that the MCMC run does not have the same inconsistancies with the NR method, and in @fig-tau-hist-both it seems that the MCMC method captures better both the galaxies with small timescales, and the galaxies with slow rising SFR's.

![$A_{del}-\text{sSFR}$ plot, where $A_{del}$ is calculated with different methods. In blue is the normalization constant as it is calculated by the MCMC method, while in red are results from Newton-Raphson.](method_comparison/sSFR_A.png){#fig-sSFR-A-both-methods}

![Distribution of $\tau$ values, from the two methods. In blue is the timescale distribution as it is calculated by the MCMC method, while in red are results from Newton-Raphson.](method_comparison/tau_hist.png){#fig-tau-hist-both}
